Follow-up Samples

### 2017-04-28 submission/manuscript.tex
complete Mary comments

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value \textless{} 0.001); however, there
was not a significant difference between those with advanced adenoma and
those with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre and post-treatment
samples for each of the diagnosis groups. Although the three models
successfully differentiated between the pre and post-treatment samples,
there was little overlap between the OTUs that were indicative of
treatment. Next, we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and carcinomas
to determine whether individuals who underwent a treatment were more
likely to have OTUs associated with normal colons. We again built Random
Forest models and measured the change in the positive probability of
having one of the three diagnoses. Only those patients that initially
had carcinomas experienced a significant decrease in positive
probability of having a lesion (P-value \textless{} 0.05). Finally, we
tested whether the type of treatment impacted the microbiota of those
diagnosed with carcinomas and were unable to detect any significant
differences in characteristics of these communities between individuals
treated with surgery alone and those treated with chemotherapy or
chemotherapy and radiation (P-value \textgreater{} 0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, thanks to improvements in screening {[}1{]}. Despite this
improvement, there are still approximately 50,000 CRC-related deaths per
year {[}2{]}. Current estimates indicate that 20-30\% of those who
undergo treatment will experience recurrence and 30-50\% of those
patients will die {[}3,4{]}. Identification of methods to assess
patients' risk of recurrence is of great importance to reduce mortality
and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}5--9{]}. Furthermore, studies that alter the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces has shown that varying community compositions can
result in varied tumor burden {[}10--12{]}. Collectively, these studies
support the hypothesis that the microbiota can alter the amount of
inflammation in the colon and with it the rate of tumorigenesis
{[}13{]}.

Building upon this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. Gut bacteria have a role in tumorigenesis and
there is promise that these populations may be useful biomarkers for
aiding in the early detection of disease.

@@

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs to limit the
likelihood that the models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
Interestingly, the 10 OTUs that were used for each model had little
overlap with each other {[}Figure 2{]}. These results support the
earlier community-wide analysis where we observed that the treatment had
an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma change
towards a microbiota associated with normal}} Next, we determined
whether treatment changed the microbiota in a way that the
post-treatment communities resembled that of patients with normal
colons. To test this, we used an expanded cohort of 423 individuals that
were diagnosed under the same protocol as having normal colons or colons
with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
constructed Random Forest models to classify the pre and post-treatment
samples as having their original diagnosis or having a normal colon. The
models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.90; Figure S2). The OTUs that
were incorporated into the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
3B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was no
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 4{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased with treatment. In fact, 6 of the 26 patients
(23.08\%) that were initially diagnosed with a carcinoma had a higher
positive probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure (P=
1). Finally, the change in the positive probability was not
significantly different between the two treatment groups (P=1). Due to
the relatively small number of samples in each treatment group, it was
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota.

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that the microbiota contains biomarkers
to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
comparing the microbiota of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients we observed changes in their microbiota. The microbiota of
patients diagnosed with carcinoma changed significantly more than the
other groups. This change resulted in communities that more closely
resembled those of patients with a normal colon. For the patients
diagnosed with carcinomas there was a larger change between their
initial and follow up samples than was observed for patients diagnosed
with either adenoma or advanced adenoma. The changes in the community
structure of those diagnosed with carcinoma were similar and resulted in
communities that more closely resembled those found in patients with
normal colons.

Understanding the effect of treatment on the microbiota of those
initially diagnosed with carcinomas may have important implications into
understanding disease recurrence. The microbiota changes associated with
treatment among patients initially diagnosed with carcinomas resulted in
community structures that resembled patients with normal colons. This
would suggest that treatment for carcinoma is not only successful for
removing the carcinoma but also at removing bacteria associated with
them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
an elevated probability of having carcinomas after treatment. This leads
us to hypothesize that these individuals may be at a high risk of
recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
probability from our study is within the expected rate of recurrence
(20-30\% {[}3,4{]}). Unfortunately, because of the design of our study,
it was not possible to determine whether individuals who had an
increased positive probability for carcinoma experienced recurrence.
Regardless, it is intriguing that it may be possible to use
microbiome-based biomarkers to not only predict the presence of lesions,
but to also assess the risk of developing them.

It is interesting that those patients diagnosed with adenoma and
advanced adenoma did not experience a shift towards a community
structure that resembled those with normal colons. This may be due to
the fundamental differences between the features of adenomas and
advanced adenomas and carcinoma. Specifically, carcinomas may create an
inflammatory milieu that would impact the structure of the community and
removal of that stimulus would alter said structure. In addition, it is
possible that the difference between the microbiota of patients with
adenoma and advanced adenoma and those with normal colons is subtle.
This is supported by the reduced ability of our models to correctly
classify patients with adenomas and advanced adenomas relative to those
diagnosed with carcinomas {[}Figure S2{]}. Given the patchy distribution
of microbiota across patients in the different diagnosis groups, it is
possible that we lacked the statistical power to adequately characterize
the change in the communities following treatment. A final hypothesis is
that the specific type of treatment altered the structure of the
microbiome. The treatment to remove adenomas and advanced adenomas was
either polyp removal or surgical resection whereas it was surgical
resection alone or in combination with chemotherapy or with chemotherapy
and radiation for individuals with carcinoma. Because chemotherapy and
radiation target rapidly growing cells, these treatments would be more
likely to cause a turn over of the colonic epithelium driving a more
significant change in the structure of the microbiota. Although, we were
able to test for an effect across these specific types of treatment, the
number of patients in each treatment group was relatively small.

It was not clear why 6 of the 26 patients with carcinoma had a higher
probability of having carcinomas after treatment. We hypothesized that
these individuals may have had more severe tumors; however, the tumor
severity of these 6 individuals (3 with Stage II and 3 with Stage III)
was similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

This study builds upon existing work that has established a role for the
microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
The most exciting future direction from the current study is the
possibility that markers within the microbiota could be used to evaluate
the effect of treatment and predict recurrence for those initially
diagnosed with carcinoma. If such an approach is effective, it might be
possible to even target the microbiota as part of combined therapies
including currently standard approaches. Our data provides additional
evidence on the importance of the microbiota in tumorigenesis by
addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}28{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of biopsies. Patients were classified as having
advanced adenoma if they had an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology. This
study was approved by the University of Michigan Institutional Review
Board. All study participants provided informed consent and the study
itself conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}29{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}30{]}, screening for
chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}32{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}33{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals {[}Table 1{]}. There were a total of 239 individuals in the
adenoma model, 262 individuals in the advanced adenoma model, and 266
indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
applied to our 67-person cohort testing prediction of adenoma intial
(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
carcinoma follow up (carcinoma n = 1).

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value \textless{} 0.001). There was no
significant change in the microbiota between patients with adenoma and
advanced adenoma, or between patients with advanced adenoma and
carcinoma (P-value \textgreater{} 0.05). Although treatment was
associated with intrapersonal changes, the change in the abundance of
individual OTUs to treatment was not consistent within diagnosis groups
(P-value \textgreater{} 0.05). Because the distribution of OTUs across
patients and diagnosis groups was irregular, we used the Random Forest
machine learning algorithm to identify groups of OTUs that allowed us to
successfully distinguish between pre and post-treatment samples for each
of the diagnosis groups. Although the three models successfully
differentiated between the pre and post-treatment samples, there was
little overlap between the OTUs that were indicative of treatment. Next,
we used a larger cohort that contained individuals with normal colons
and those with adenomas, advanced adenomas, and carcinomas to determine
whether individuals who underwent treatment were more likely to have
OTUs associated with normal colons. We again built Random Forest models
and measured the change in the positive probability of having one of the
three diagnoses. Only patients who had carcinomas experienced a
significant decrease in positive probability of having a lesion after
treatment (P-value \textless{} 0.05), indicating that the microbial
milieu of the colon more closely resembled that of a normal colon.
Finally, we tested whether the type of treatment impacted the microbiota
of those diagnosed with carcinomas and were unable to detect any
significant differences in characteristics of these communities between
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, predominately due to improvements in screening {[}2{]}.
Despite this improvement, there are still approximately 50,000
CRC-related deaths per year {[}1{]}. Current estimates indicate that
20-30\% of those who undergo treatment will experience recurrence and
35\% of all patients will die within five years {[}3--5{]}.
Identification of methods to assess patients' risk of recurrence is of
great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}6--10{]}. Furthermore, studies that shift the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces have shown that varying community compositions can
result in varied tumor burden {[}11--13{]}. Collectively, these studies
support the hypothesis that the microbiota can alter the amount of
inflammation in the colon and with it the rate of tumorigenesis
{[}14{]}.

Building upon this evidence, several human studies have identified
unique signatures of colonic lesions {[}15--20{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. This suggests that gut bacteria have a role in
tumorigenesis with potential as useful biomarkers for aiding in the
early detection of disease.

@@

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the
irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. Because of the relatively small number of subjects in
each group, we restricted our models to only incorporate 10 OTUs to
limit the likelihood that the models would overfit the data. Despite
this restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
Interestingly, the 10 OTUs that were used for each model had little
overlap with each other {[}Figure 2{]}. These results support the
earlier community-wide analysis where we observed that the treatment had
an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.90; Figure S2). The OTUs that
were incorporated into the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
3B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was no
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 4{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased with treatment. Only, 6 of the 26 patients
(23.08\%) who were diagnosed with a carcinoma had a higher positive
probability after treatment; one of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that, although
there were changes in the microbiota associated with treatment, those
experienced by patients with carcinoma after treatment yielded gut
bacterial communities of greater similarity to that of a normal colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure
(P-value = 0.298). Finally, the change in the positive probability was
not significantly different between the two treatment groups (P-value =
0.999). Due to the relatively small number of samples in each treatment
group, it was difficult to make a definitive statement regarding the
specific type of treatment on the amount of change in the structure of
the microbiota.

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that microbiome-derived biomarkers can
detect colonic lesions {[}6--10,24--28{]}. Our study focused on
comparing the microbiota of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients, we observed changes in their microbiota. After treatment,
the microbiota of patients with carcinoma changed significantly more
than the other groups. This change resulted in communities that more
closely resembled those of patients with a normal colon. This may
suggest that treatment for carcinoma is not only successful for removing
the carcinoma but also at reducing the associated bacterial communities.
Understanding the effect of treatment on the microbiota of those
diagnosed with carcinomas may have important implications for reducing
disease recurrence. It is intriguing that it may be possible to use
microbiome-based biomarkers to not only predict the presence of lesions
but to assess the risk of recurrence.

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. In addition, it is possible that the difference
between the microbiota of patients with adenoma and advanced adenoma and
those with normal colons is subtle. This is supported by the reduced
ability of our models to correctly classify patients with adenomas and
advanced adenomas relative to those diagnosed with carcinomas {[}Figure
S2{]}. Given the irregular distribution of microbiota across patients in
the different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the
structure of the microbiome. The treatment to remove adenomas and
advanced adenomas was either polyp removal or surgical resection whereas
it was surgical resection alone or in combination with chemotherapy or
with chemotherapy and radiation for individuals with carcinoma. Because
chemotherapy and radiation target rapidly growing cells, these
treatments would be more likely to cause a turnover of the colonic
epithelium driving a more significant change in the structure of the
microbiota. Although, we were able to test for an effect across these
specific types of treatment, the number of patients in each treatment
group was relatively small.

This study expands upon existing research that has established a role
for the microbiota in tumorigenesis and that demonstrated the utility of
microbiome-based biomarkers to predict the presence of colonic lesions.
The most exciting future direction from the current study is the
possibility that markers within the microbiota could be used to evaluate
the effect of treatment and predict recurrence for those diagnosed with
carcinoma. If such an approach is effective, it might be possible to
target the microbiota as part of adjuvant therapy. Our data provides
additional evidence on the importance of the microbiota in tumorigenesis
by addressing the recovery of the microbiota after treatment and opens
interesting avenues of research into how these changes may affect
recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period {[}29{]}. Our study cohort consisted
of 67 individuals with an initial sample as described and a follow up
sample obtained between 188 - 546 days after treatment of lesion
{[}Table 1{]}. Patients were diagnosed by colonoscopic examination and
histopathological review of biopsies. Patients were classified as having
advanced adenoma if they had an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology. This
study was approved by the University of Michigan Institutional Review
Board. All study participants provided informed consent and the study
itself conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}30{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}31{]}, screening for
chimeras using UCHIME {[}32{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}33{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}34{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}35{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, the advanced adenoma
model classified normal versus advanced adenoma, and the carcinoma model
classified normal versus carcinoma. The total number of individuals in
the treatment models was 67 individuals. There were a total of 22
individuals in the treatment adenoma model, 19 individuals in the
treatment advanced adenoma model, and 26 individuals in the treatment
carcinoma model {[}Table 1{]}. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the adenoma model, 262 individuals in
the advanced adenoma model, and 266 individuals in the carcinoma model
{[}Table 2{]}. Each model was then applied to our 67-person cohort
{[}Table 1{]} which assesed the prediction of pre-treatment adenoma
(adenoma n = 22) versus post-treatment adenoma (adenoma n = 0),
pre-treatment advanced adenoma (advanced adenoma n = 19) versus
post-treatment advanced adenoma (advanced adenoma n = 0), pre-treatment
carcinoma (carinoma n = 26) versus post-treatment carcinoma (carcinoma n
= 1).
